---
title: "Assignment 5"
author: "Yesdi Christian Calvin, Aulia Dini"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

This is the Github link of the project : https://github.com/aulia0716/Yesdi-Aulia-A5

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You may work in pairs or individually for this assignment. Make sure you
join a group in Canvas if you are working in pairs. Turn in this
assignment as an HTML or PDF file to ELMS. Make sure to include the R
Markdown or Quarto file that was used to generate it. Include the GitHub
link for the repository containing these files.

```{r, warning=FALSE, message=FALSE}
library(censusapi)
library(tidyverse)
library(magrittr)
library(factoextra)
```

## Exploring ACS Data

In this notebook, we use the Census API to gather data from the American
Community Survey (ACS). This requires an access key, which can be
obtained here:

https://api.census.gov/data/key_signup.html

```{r}
cs_key <- "7b7f40a6d561e9e43edb14fec7e8b645055a65ed"

```


```{r}
acs_il_c <- getCensus(name = "acs/acs5",
                  vintage = 2016,
                  vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"),
                  region = "county:*",
                  regionin = "state:17",
                  key = cs_key) %>%
                  rename(pop = B01003_001E, 
                  hh_income = B19013_001E, 
                  income = B19301_001E)
head(acs_il_c)
```

Pull map data for Illinois into a data frame.

```{r}
il_map <- map_data("county", region = "illinois")
head(il_map)
```



Join the ACS data with the map data. Note that `il_map` has a column
`subregion` which includes county names. We need a corresponding
variable in the ACS data to join both data sets. This needs some
transformations, among which the function `tolower()` might be useful.
Call the joined data `acs_map`.

```{r}
# clean name of location in the acs_il data
acs_il_c <- acs_il_c %>%
# remove all words after certain character
mutate(county = gsub("(.*),\\s*(.*?)\\s*$", "\\1", NAME))
# remove the last word
acs_il_c$county <- sub("\\s+\\w+$", "", acs_il_c$county)
# check the result
head(acs_il_c)
```


```{r}
# First, let's ensure that county names in both datasets are in lower case to have a uniform format.

# Convert county names in il_map to lower case.
il_map$subregion <- tolower(il_map$subregion)

# Assume 'county' is the column in your ACS data containing county names. Make it lower case.
acs_il_c$county <- tolower(acs_il_c$county)

# After checking the names of counties in the two datasets manually, we found the four counties below whose names are different between these two datasets. Therefore, we modify the names to make them identical and match between the two datasets.

acs_il_c <- acs_il_c %>%
  mutate(
    county = case_when(
      county == "dekalb" ~ "de kalb",
      county == "dupage" ~ "du page",
      county == "lasalle" ~ "la salle",
      county == "st. clair" ~ "st clair",
      TRUE ~ county  # Keep the original name if no match
    )
  )

```


```{r}
# We join the two datasets using left_join
library(dplyr)

acs_map <- left_join(acs_il_c, il_map, by = c("county" = "subregion"))  

```


After you do this, plot a map of Illinois with Counties colored by per
capita income.

```{r}
#| fig.height = 6.5, fig.width = 6
library(ggplot2)

ggplot(acs_map) + 
  geom_polygon(aes(x = long, y = lat, group = group, fill = income))

```

## Hierarchical Clustering

We want to find clusters of counties that are similar in their population, average household income and per capita income. First, clean the data so that you have the appropriate variables to use for clustering. Next, create the distance matrix of the cleaned data. This distance matrix can be used to cluster counties, e.g. using the ward method.

```{r}
#STEPS
#1. clean the data so that we have the appropriate variables to use for clustering
#2. create the distance matrix of the cleaned data. This distance matrix can be used to cluster counties.


#1
library(cluster)

# Filter and clean dataset
cluster_df <- acs_map %>%
              select(county, pop, hh_income, income) %>%
              na.omit() %>%
              # Convert all columns to numeric type
              mutate(across(c(pop, hh_income, income), as.numeric)) 

# Scale the variables
cluster_df_scaled <- cluster_df
# Scale all columns except the first one (County Names)
cluster_df_scaled[,-1] <- scale(cluster_df_scaled[,-1])  


#2
# Compute the distance matrix
dist_mat <- dist(cluster_df_scaled[,-1])

# Hierarchical clustering using Ward's Method
hc <- hclust(dist_mat, method = "ward.D2")  # Apply the Ward Method

# Display a dendrogram of the result
plot(hc, labels = cluster_df$county, hang = -1,
     main = "Dendrogram of Counties in Illinois",
     sub = "",
     ylab = "Height (Ward)")
```


Plot the dendrogram to find a reasonable number of clusters. Draw boxes
around the clusters of your cluster solution.

```{r}
# Create dendogram
plot(hc, main = "Ward", xlab = "", sub = "")

# Draw boxes around the cluster solution. Based on our observation from the dendrogram, we inspect that the number of clusters is 6, so we use it as the k value in this case.

rect.hclust(hc, 
            k = 6, 
            border = "red")
```

Visualize the county clusters on a map. For this task, create a new `acs_map` object that now also includes cluster membership as a new column. This column should be called `cluster`.

```{r, warning=FALSE}
# Cut the dendrogram at a specific height to determine the number of clusters
cut_height <- 17.5  # We adjust this based on our visual inspection of the dendrogram
cluster_memberships <- cutree(hc, h = cut_height)

# Add cluster membership to the acs_map dataset
acs_map_with_clusters <- acs_map %>%
  mutate(cluster = as.factor(cluster_memberships))

# Print the first few rows of the new acs_map object to verify the addition of the 'cluster' column
head(acs_map_with_clusters)

# Map with color indicating cluster
map_with_clusters <- ggplot() +
  geom_polygon(data = acs_map_with_clusters, aes(x = long, y = lat, group = group, fill = cluster), color = "blue", size = 0.2) +
  coord_map() +
  scale_fill_discrete(name = "Cluster Number") +
  theme_minimal() +
  labs(title = "Clustered Counties in Illinois",
       x = "Longitude",
       y = "Latitude")

# Print the map
print(map_with_clusters)

```

## Census Tracts

For the next section we need ACS data on a census tract level. We use
the same variables as before.

```{r}
acs_il_t <- getCensus(name = "acs/acs5",
                     vintage = 2016, 
                     vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"),
                     region = "tract:*", 
                     regionin = "state:17",
                     key = cs_key) %>%
             mutate_all(list(~ ifelse(.==-666666666, NA, .))) %>%
             rename(pop = B01003_001E, 
                    hh_income = B19013_001E, 
                    income = B19301_001E)
head(acs_il_t)
```

## k-Means

As before, clean our data for clustering census tracts based on
population, average household income and per capita income.

```{r}
cluster_ct <- acs_il_t %>%
  select(county, pop, hh_income, income) %>%
  mutate(pop = as.numeric(pop), 
         hh_income = as.numeric(hh_income),
         income = as.numeric(income)) %>%
  na.omit() # omit rows with any remaining NA values

head(cluster_ct)
```

```{r}
# subset the data
c_data2 <-
  cluster_ct %>%
   select(county, pop, hh_income, income) %>%
  mutate(pop = as.numeric(pop), 
         hh_income = as.numeric(hh_income),
         income = as.numeric(income)) %>%
  na.omit()
```

Since we want to use K Means in this section, we start by determining the optimal number of K that results in Clusters with low within but high between variation. Plot within cluster sums of squares for a range of K (e.g. up to 20).

```{r, warning=FALSE}
# check the optimum cluster
fviz_nbclust(c_data2, #data set we want to use
             kmeans, #cluster method
             method = "wss", #method used for estimating the optimal number of clusters
             k.max = 20)
```



Run `kmeans()` for the optimal number of clusters based on the plot
above.

```{r}
# we inspect that the optimal number of clusters is 6 since it is the elbow of the graph
optimal_k <- 6

# Run k-means clustering 
set.seed(123) # For reproducibility
km_12 <- kmeans(c_data2, centers = optimal_k, nstart = 50)

print(km_12)
```


Find the mean population, household income and per capita income grouped
by clusters. In addition, display the most frequent county that can be
observed within each cluster.

```{r}

c_data2$cluster <- km_12$cluster


cluster_summary <- c_data2 %>%
  group_by(cluster) %>%
  summarise(
    mean_pop = mean(pop, na.rm = TRUE),
    mean_hh_income = mean(hh_income, na.rm = TRUE),
    mean_income = mean(income, na.rm = TRUE),
    most_frequent_county = names(which.max(table(county)))
  )

print(cluster_summary)
```

As you might have seen earlier, it's not always clear which number of
clusters is the optimal choice. To automate K Means clustering, program
a function based on `kmeans()` that takes K as an argument. You can fix
the other arguments, e.g. such that a specific dataset is always used
when calling the function.

```{r}
# define function
kmeans_custom <- function(k, data = c_data2) {
  set.seed(123)
  result_cluster <- kmeans(c_data2, centers = k, nstart = 50)
  return(result_cluster)
}

```


We want to utilize this function to iterate over multiple Ks (e.g., K =
2, ..., 10) and \-- each time \-- add the resulting cluster membership
as a new variable to our (cleaned) original data frame (`acs_il_t`).
There are multiple solutions for this task, e.g. think about the `apply`
family or `for` loops.

```{r}
# Make a copy of your data to avoid changing the original
acs_il_t_clustered <- c_data2

# Iteratively perform k-means clustering for K = 2 to 10
for (k in 2:10) {
  # Run k-means and store the result
  kmeans_result <- kmeans_custom(k)
  
  # Add cluster membership to the dataframe
  acs_il_t_clustered[paste("cluster_k", k, sep="_")] <- kmeans_result$cluster
}
```


Finally, display the first rows of the updated data set (with multiple
cluster columns).

```{r}
head(acs_il_t_clustered)
```

